---
title: "Predictors of In-Hospital Mortality — GEMINI Cohort"
author: "Reference Answer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: kate
    df_print: kable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  comment    = "",
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5
)

pacman::p_load(
  tidyverse, lubridate, table1, broom,
  janitor, knitr, kableExtra, stringr,
  pROC, car, scales, pheatmap, RColorBrewer
)

theme_set(theme_minimal(base_size = 12))
```

---

# 1 — Cohort Construction

**Question:** *Who is in the study, and how did we arrive at the analytic sample?*

**Why:** Every observational study must define a reproducible cohort with
explicit, sequential inclusion/exclusion criteria. The attrition table lets
reviewers assess generalizability, detect selection bias, and understand
exactly which records were dropped and why.

```{r load-data}
# ---- Admin/encounter-level file ----
ipadmdad_raw <- read.csv("ipadmdad.csv") %>%
  mutate(
    admission_date_time = ymd_hm(admission_date_time),
    discharge_date_time = ymd_hm(discharge_date_time)
  )
```

```{r cohort-build}
n0 <- nrow(ipadmdad_raw)

# Step 1 — Adults only (age >= 18, non-missing)
cohort <- ipadmdad_raw %>%
  filter(!is.na(age), age >= 18)
n1 <- nrow(cohort)

# Step 2 — Non-missing discharge disposition (needed to derive outcome)
cohort <- cohort %>%
  filter(!is.na(discharge_disposition),
         trimws(as.character(discharge_disposition)) != "")
n2 <- nrow(cohort)

# Step 3 — LOS >= 4 hours (removes miscodings, rapid transfers)
cohort <- cohort %>%
  mutate(los_hrs = as.numeric(
    difftime(discharge_date_time, admission_date_time, units = "hours")
  )) %>%
  filter(!is.na(los_hrs), los_hrs >= 4)
n3 <- nrow(cohort)

# Step 4 — Study window: 3 Ontario fiscal years (Apr 2019 – Mar 2022)
cohort <- cohort %>%
  filter(admission_date_time >= ymd("2019-04-01"),
         admission_date_time <  ymd("2022-04-01"))
n4 <- nrow(cohort)

tibble(
  Step    = c("Raw admissions",
              "Adults (age ≥ 18)",
              "Non-missing discharge disposition",
              "LOS ≥ 4 hours",
              "Within study period (Apr 2019 – Mar 2022)"),
  N       = c(n0, n1, n2, n3, n4),
  Excluded = c(NA_integer_, n0 - n1, n1 - n2, n2 - n3, n3 - n4)
) %>%
  kable(caption = "Table S1 — Cohort attrition",
        format.args = list(big.mark = ",")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

The analytic cohort includes **`r format(n4, big.mark = ",")`** encounters
across **`r n_distinct(cohort$hospital_num)`** hospitals.

---

# 2 — Data Quality & Sanity Checks

**Question:** *Is the data plausible and complete enough to support the planned
analysis?*

**Why:** EHR data is messy. Before any modelling, you must verify that key
variables have realistic distributions, that missingness isn't systematic (e.g.,
concentrated in specific hospitals), and that obvious data errors are flagged.
This is 30% of the GEMINI data scientist role and the step most candidates skip.

## 2.1 Plausible-range checks on continuous variables

```{r sanity-ranges}
tibble(
  Variable = c("Age", "LOS (hours)"),
  Min      = c(min(cohort$age), min(cohort$los_hrs)),
  Q1       = c(quantile(cohort$age, 0.25), quantile(cohort$los_hrs, 0.25)),
  Median   = c(median(cohort$age), median(cohort$los_hrs)),
  Q3       = c(quantile(cohort$age, 0.75), quantile(cohort$los_hrs, 0.75)),
  Max      = c(max(cohort$age), max(cohort$los_hrs)),
  `N flagged` = c(
    sum(cohort$age > 110),                # implausible age
    sum(cohort$los_hrs > 365 * 24)        # > 1 year
  )
) %>%
  mutate(across(Min:Max, ~ round(.x, 1))) %>%
  kable(caption = "Sanity check — continuous variable ranges") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

## 2.2 Age distribution by hospital

```{r age-distribution, fig.height=4}
ggplot(cohort, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#457B9D", colour = "white", linewidth = 0.2) +
  facet_wrap(~ hospital_num, scales = "free_y", ncol = 4) +
  labs(x = "Age (years)", y = "Count",
       title = "Age distribution by hospital",
       subtitle = "Check for implausible spikes, truncation, or rounding artefacts") +
  theme(strip.text = element_text(size = 8))
```

## 2.3 Missingness heatmap by hospital

```{r missingness-heatmap, fig.height=5, fig.width=9}
# Compute % missing per hospital for all columns of interest
cols_check <- c("gender", "hospital_num")
# Add any other columns from your data that were NOT used as exclusion criteria
# age, discharge_disposition, admission/discharge times are guaranteed non-missing
# because you already filtered on them in Section 1

miss_by_hosp <- cohort %>%
  group_by(hospital_num) %>%
  summarise(across(all_of(cols_check), ~ round(100 * mean(is.na(.x)), 2)),
            .groups = "drop") %>%
  column_to_rownames("hospital_num")
mat <- as.matrix(miss_by_hosp)
if (diff(range(mat, na.rm = TRUE)) < 0.001) {
  cat("Missingness is uniformly", round(mat[1,1], 1),
      "% across all hospitals — no heatmap needed.")
} else {
  pheatmap(
    mat,
    cluster_rows   = TRUE,
    cluster_cols    = FALSE,
    display_numbers = TRUE,
    number_format   = "%.1f",
    color           = colorRampPalette(c("white", "#E63946"))(50),
    main            = "% Missing by Hospital — Core Admin Variables",
    fontsize        = 10,
    angle_col       = 45
  )
}
```

*Interpretation: If missingness clusters in specific hospitals, it may indicate
systematic data-extraction issues rather than random missingness — this affects
how we handle incomplete records downstream.*

## 2.4 Crude mortality rate by hospital (face-validity check)

```{r mortality-by-hospital, fig.height=4}
hosp_mort <- cohort %>%
  group_by(hospital_num) %>%
  summarise(n = n(), deaths = sum(died <- as.integer(
    discharge_disposition %in% c(7, 72, 73))),
    rate = deaths / n, .groups = "drop") %>%
  mutate(hospital_num = reorder(factor(hospital_num), rate))

ggplot(hosp_mort, aes(x = hospital_num, y = rate)) +
  geom_col(fill = "#264653") +
  geom_hline(yintercept = mean(hosp_mort$rate), linetype = "dashed",
             colour = "#E76F51", linewidth = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(x = "Hospital", y = "Crude mortality rate",
       title = "Crude in-hospital mortality by site",
       subtitle = "Dashed line = overall mean. Large outliers may reflect case-mix or data issues.") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

---

# 3 — Feature Engineering

**Question:** *What analytic variables do we need to derive from raw fields, and
how are they defined?*

**Why:** Raw EHR data is never analysis-ready. Outcomes must be coded from
administrative codes, comorbidities extracted via ICD-10-CA regex, and temporal
features (LOS categories, admission timing) created explicitly. Every derivation
must be documented to ensure reproducibility.

## 3.1 Outcome — in-hospital mortality

```{r derive-outcome}
# CIHI discharge disposition codes: 7, 72, 73 = death
cohort <- cohort %>%
  mutate(died = as.integer(discharge_disposition %in% c(7, 72, 73)))

cat("Deaths:", sum(cohort$died),
    sprintf("(%.1f%%)\n", 100 * mean(cohort$died)))
cat("Survivors:", sum(cohort$died == 0),
    sprintf("(%.1f%%)", 100 * mean(cohort$died == 0)))
```

## 3.2 Length of stay

```{r derive-los}
cohort <- cohort %>%
  mutate(
    los_days     = los_hrs / 24,
    los_days_cat = cut(los_days,
                       breaks = c(0, 3, 7, 14, Inf),
                       labels = c("≤3d", "4–7d", "8–14d", ">14d"),
                       right  = TRUE)
  )
```

## 3.3 Comorbidity flags (ICD-10-CA)

We derive three common comorbidity indicators from admission diagnoses
(types M, 1, 2, 3). This is a **simplified burden count**, not a validated
Charlson or Elixhauser index.

```{r derive-comorbidities}
ipdiagnosis_raw <- read.csv("ipdiagnosis.csv")

admit_dx <- ipdiagnosis_raw %>%
  filter(genc_id %in% cohort$genc_id,
         diagnosis_type %in% c("M", "1", "2", "3"),
         !is.na(diagnosis_code), trimws(diagnosis_code) != "")

comorb <- admit_dx %>%
  mutate(
    diabetes = as.integer(str_starts(diagnosis_code, "E1[01]")),
    renal    = as.integer(str_starts(diagnosis_code, "N1[78]")),
    cancer   = as.integer(str_starts(diagnosis_code, "C"))
  ) %>%
  group_by(genc_id) %>%
  summarise(diabetes = max(diabetes),
            renal    = max(renal),
            cancer   = max(cancer),
            n_dx     = n_distinct(diagnosis_code),
            .groups  = "drop")

cohort <- cohort %>%
  left_join(comorb, by = "genc_id") %>%
  mutate(across(c(diabetes, renal, cancer), ~ replace_na(.x, 0L)),
         n_dx      = replace_na(n_dx, 0L),
         dx_burden = diabetes + renal + cancer)
```

## 3.4 Admission timing

```{r derive-timing}
cohort <- cohort %>%
  mutate(
    weekend_admit = as.integer(wday(admission_date_time) %in% c(1, 7)),
    fiscal_year   = if_else(month(admission_date_time) >= 4,
                            year(admission_date_time),
                            year(admission_date_time) - 1L),
    admit_month   = floor_date(admission_date_time, "month")
  )
```

## 3.5 Early lab values (first 24 hours)

```{r derive-labs}
lab_raw <- read.csv("lab.csv")

lab_early <- lab_raw %>%
  filter(genc_id %in% cohort$genc_id) %>%
  inner_join(cohort %>% select(genc_id, admission_date_time), by = "genc_id") %>%
  mutate(
    collection_date_time = ymd_hm(collection_date_time),
    hrs_since_admit = as.numeric(
      difftime(collection_date_time, admission_date_time, units = "hours"))
  ) %>%
  filter(hrs_since_admit >= 0, hrs_since_admit <= 24)

# First value per test per encounter
lab_first <- lab_early %>%
  group_by(genc_id, test_type_mapped_omop) %>%
  slice_min(collection_date_time, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(genc_id, test_type_mapped_omop, result_value)

# Pivot to wide: one column per lab test
lab_wide <- lab_first %>%
  pivot_wider(names_from  = test_type_mapped_omop,
              values_from = result_value)

cohort <- cohort %>%
  left_join(lab_wide, by = "genc_id")

# Identify lab columns dynamically
lab_cols <- setdiff(names(lab_wide), "genc_id")
cat("Lab tests available:", paste(lab_cols, collapse = ", "))
```

### Lab missingness assessment

```{r lab-missingness, fig.height=5, fig.width=9}
lab_miss_hosp <- cohort %>%
  group_by(hospital_num) %>%
  summarise(across(all_of(lab_cols), ~ round(100 * mean(is.na(.x)), 1)),
            .groups = "drop") %>%
  column_to_rownames("hospital_num")

pheatmap(
  as.matrix(lab_miss_hosp),
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  display_numbers = TRUE,
  number_format   = "%.0f",
  color           = colorRampPalette(c("#F1FAEE", "#A8DADC", "#E63946"))(50),
  main            = "% Missing Lab Values (First 24h) by Hospital",
  fontsize        = 9,
  angle_col       = 45
)
```

*Systematic missingness in specific hospitals suggests data-extraction gaps rather
than clinical absence — these labs were likely collected but not captured.*

### Lab plausible-range check

```{r lab-ranges}
# Quick sanity on any numeric lab columns
cohort %>%
  summarise(across(all_of(lab_cols),
    list(
      n_avail = ~ sum(!is.na(.x)),
      median  = ~ round(median(.x, na.rm = TRUE), 1),
      p5      = ~ round(quantile(.x, 0.05, na.rm = TRUE), 1),
      p95     = ~ round(quantile(.x, 0.95, na.rm = TRUE), 1)
    ), .names = "{.col}||{.fn}")) %>%
  pivot_longer(everything(),
               names_to = c("Lab", ".value"),
               names_sep = "\\|\\|") %>%
  kable(caption = "Lab value distributions (first 24h)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

---

# 4 — Descriptive & Exploratory Analysis

**Question:** *How do patient characteristics differ between those who died and
those who survived? Are there temporal patterns or non-linear relationships we
should account for in modelling?*

**Why:** Table 1 is mandatory in every clinical manuscript — it establishes
face-validity and identifies potential confounders. Exploratory plots go further:
they reveal non-linearity, temporal trends (e.g., COVID waves), and unexpected
patterns that inform model specification.

## 4.1 Table 1 — Characteristics by outcome

```{r table1}
cohort_tbl <- cohort %>%
  mutate(
    Outcome  = factor(died, 0:1, c("Survived", "Died")),
    Sex      = factor(gender, c("M", "F"), c("Male", "Female")),
    Diabetes = factor(diabetes, 0:1, c("No", "Yes")),
    CKD      = factor(renal, 0:1, c("No", "Yes")),
    Cancer   = factor(cancer, 0:1, c("No", "Yes")),
    `Diagnosis burden` = factor(n_dx),
    `LOS category` = los_days_cat,
    `Weekend admission` = factor(weekend_admit, 0:1, c("No", "Yes"))
  )

# table1 produces stratified descriptive stats with p-values via built-in tests
table1(~ age + Sex + los_days + `LOS category` + Diabetes + CKD + Cancer +
         `Weekend admission` + n_dx | Outcome,
       data    = cohort_tbl,
       caption = "Table 1 — Patient characteristics by in-hospital mortality")
```

## 4.2 Monthly mortality trend

```{r monthly-trend, fig.height=4}
monthly <- cohort %>%
  group_by(admit_month) %>%
  summarise(n = n(), deaths = sum(died),
            rate = deaths / n, .groups = "drop")

ggplot(monthly, aes(x = admit_month, y = rate)) +
  geom_line(colour = "#264653", linewidth = 0.7) +
  geom_point(colour = "#264653", size = 1.5) +
  geom_smooth(method = "loess", se = TRUE, colour = "#E76F51",
              fill = "#E76F51", alpha = 0.15, linewidth = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "3 months") +
  labs(x = NULL, y = "Monthly mortality rate",
       title = "In-hospital mortality over time",
       subtitle = "Loess smoother highlights pandemic-period trends") +
  theme(axis.text.x = element_text(size = 8))
```

*Look for mortality spikes during COVID-19 wave 1 (spring 2020) and wave 2
(winter 2020–21). This motivates adjusting for fiscal year in the model.*

## 4.3 Age–mortality relationship (non-linearity check)

```{r age-mortality-curve, fig.height=4}
# Binned approach: compute mortality rate per 5-year age bin
age_bins <- cohort %>%
  mutate(age_bin = cut(age, breaks = seq(18, 110, by = 5), right = FALSE)) %>%
  group_by(age_bin) %>%
  summarise(n = n(), rate = mean(died), .groups = "drop") %>%
  filter(!is.na(age_bin)) %>%
  mutate(mid = seq(20, by = 5, length.out = n()))

ggplot(age_bins, aes(x = mid, y = rate)) +
  geom_point(aes(size = n), colour = "#264653", alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, colour = "#E76F51",
              fill = "#E76F51", alpha = 0.15) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_size_continuous(name = "N encounters", labels = comma) +
  labs(x = "Age (5-year midpoint)", y = "Mortality rate",
       title = "Mortality by age group",
       subtitle = "Non-linear uptick above ~75 suggests spline or polynomial may be warranted")
```

## 4.4 LOS distribution by outcome

```{r los-violin, fig.height=4}
ggplot(cohort, aes(x = factor(died, 0:1, c("Survived", "Died")),
                   y = los_days)) +
  geom_violin(aes(fill = factor(died)), alpha = 0.5, show.legend = FALSE) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  scale_fill_manual(values = c("#457B9D", "#E63946")) +
  scale_y_log10(labels = comma) +
  labs(x = NULL, y = "Length of stay (days, log scale)",
       title = "Length of stay distribution by outcome") +
  annotation_logticks(sides = "l")
```

## 4.5 Lab correlation heatmap

```{r lab-corr-heatmap, fig.height=6, fig.width=7}
# Pairwise Spearman correlations among available lab values
lab_mat <- cohort %>%
  select(all_of(lab_cols)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")

pheatmap(
  lab_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(
    rev(brewer.pal(9, "RdBu")))(100),
  breaks          = seq(-1, 1, length.out = 101),
  main            = "Spearman correlation — First-24h lab values",
  fontsize        = 9
)
```

*Strong correlations between labs (e.g., creatinine and urea) signal potential
multicollinearity if both are included in the regression.*

## 4.6 Comorbidity co-occurrence

```{r comorb-co-occurrence, fig.height=3.5, fig.width=5}
comorb_mat <- cohort %>%
  select(diabetes, renal, cancer) %>%
  cor(method = "phi") # binary-binary: phi = Pearson on 0/1 vars

pheatmap(
  comorb_mat,
  display_numbers = TRUE,
  number_format   = "%.2f",
  color           = colorRampPalette(c("#F1FAEE", "#1D3557"))(50),
  main            = "Comorbidity co-occurrence (φ correlation)"
)
```

---

# 5 — Statistical Modelling

**Question:** *Which patient-level factors are independently associated with
in-hospital mortality after adjusting for confounders — and does adding early
lab data improve prediction?*

**Why:** Multivariable logistic regression isolates each predictor's contribution
while holding others constant, producing adjusted ORs — the standard output in
clinical epi. Comparing a base model (demographics + comorbidities) against a
lab-augmented model quantifies the added predictive value of acute physiology.

## 5.1 Base model — demographics + comorbidities

Hospital is included as a fixed effect to absorb site-level differences in
case-mix and practice. With `r n_distinct(cohort$hospital_num)` hospitals, a
random-intercept model could be considered, but fixed effects are adequate for
our focus on patient-level predictors.

```{r base-model}
mod_base <- glm(
  died ~ age + gender + diabetes + renal + cancer +
    weekend_admit + factor(fiscal_year) + factor(hospital_num),
  data   = cohort,
  family = binomial
)
```

### Odds ratio table

```{r or-table}
tidy(mod_base, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(!str_starts(term, "factor\\(hospital_num\\)")) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 2)),
         p.value = format.pval(p.value, digits = 2, eps = 0.001)) %>%
  select(Term = term, OR = estimate,
         `2.5%` = conf.low, `97.5%` = conf.high,
         p = p.value) %>%
  kable(caption = "Table 2 — Adjusted ORs, base model (hospital effects suppressed)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

### Multicollinearity check (VIF)

```{r vif-check}
# Fit reduced model without factor(hospital_num) for clean VIF
mod_vif <- glm(
  died ~ age + gender + diabetes + renal + cancer +
    weekend_admit + factor(fiscal_year),
  data = cohort, family = binomial
)

vif_vals <- vif(mod_vif)
# For factors, GVIF^(1/(2*Df)) is the comparable metric
if (is.matrix(vif_vals)) {
  vif_tbl <- as.data.frame(vif_vals) %>%
    rownames_to_column("Term") %>%
    mutate(GVIF_adj = round(`GVIF^(1/(2*Df))`, 2)) %>%
    select(Term, GVIF_adj)
} else {
  vif_tbl <- tibble(Term = names(vif_vals), VIF = round(vif_vals, 2))
}

kable(vif_tbl, caption = "VIF — values > 5 suggest concerning collinearity") %>%
  kable_styling(full_width = FALSE)
```

### Discrimination — ROC & AUC

```{r base-auc, fig.height=5, fig.width=5}
cohort$pred_base <- predict(mod_base, type = "response")
roc_base <- roc(cohort$died, cohort$pred_base, quiet = TRUE)
auc_base <- ci.auc(roc_base)

# ROC curve
ggroc(roc_base, colour = "#264653", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey50") +
  labs(title = "ROC — Base model",
       subtitle = sprintf("AUC = %.3f (95%% CI %.3f–%.3f)",
                          auc_base[2], auc_base[1], auc_base[3]),
       x = "Specificity", y = "Sensitivity") +
  coord_equal()
```

### Calibration plot

```{r calibration-plot, fig.height=5, fig.width=5}
# Decile-based calibration
cohort <- cohort %>%
  mutate(pred_decile = ntile(pred_base, 10))

cal_df <- cohort %>%
  group_by(pred_decile) %>%
  summarise(
    mean_predicted = mean(pred_base),
    mean_observed  = mean(died),
    n = n(),
    .groups = "drop"
  )

ggplot(cal_df, aes(x = mean_predicted, y = mean_observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(size = n), colour = "#E76F51", alpha = 0.8) +
  geom_line(colour = "#E76F51", linewidth = 0.5) +
  scale_size_continuous(name = "N per decile", labels = comma) +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Mean predicted probability", y = "Observed proportion",
       title = "Calibration plot — Base model",
       subtitle = "Perfect calibration lies on the diagonal") +
  coord_equal(xlim = c(0, NA), ylim = c(0, NA))
```

### Forest plot

```{r forest-plot, fig.width=8, fig.height=5}
or_df <- tidy(mod_base, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)",
         !str_starts(term, "factor\\(hospital_num\\)")) %>%
  mutate(
    label = case_match(
      term,
      "age"                      ~ "Age (per year)",
      "genderM"                  ~ "Male sex",
      "diabetes"                 ~ "Diabetes",
      "renal"                    ~ "CKD",
      "cancer"                   ~ "Cancer",
      "weekend_admit"            ~ "Weekend admission",
      "factor(fiscal_year)2020"  ~ "FY 2020 vs 2019",
      "factor(fiscal_year)2021"  ~ "FY 2021 vs 2019"
    ),
    significant = p.value < 0.05
  )

ggplot(or_df, aes(x = estimate, y = reorder(label, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high, colour = significant),
    size = 0.6
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#D62828", "FALSE" = "#264653"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
    name   = NULL
  ) +
  labs(x = "Adjusted Odds Ratio (95% CI)", y = NULL,
       title = "Forest plot — Base model",
       caption = "Hospital fixed effects estimated but suppressed. Reference: FY 2019.") +
  theme(legend.position   = "bottom",
        panel.grid.minor  = element_blank(),
        panel.grid.major.y = element_blank())
```

## 5.2 Lab-augmented model & comparison

We add selected first-24h lab values to the base model. Encounters with missing
labs are excluded from this comparison (complete-case analysis on the overlapping
subset for fair comparison).

```{r lab-model}
# Pick 2–3 clinically meaningful labs that exist in the data
# (adjust column names based on what's available)
lab_predictors <- intersect(lab_cols, names(cohort))

# Use at most 3 labs to keep it parsimonious — pick by clinical relevance
# Example: hemoglobin, creatinine, sodium (adjust to actual column names)
if (length(lab_predictors) > 3) {
  lab_predictors <- lab_predictors[1:3]  # take first 3; replace with clinical picks
}

# Build formula dynamically
lab_formula <- as.formula(paste(
  "died ~ age + gender + diabetes + renal + cancer +",
  "weekend_admit + factor(fiscal_year) + factor(hospital_num) +",
  paste(lab_predictors, collapse = " + ")
))

# Complete-case subset for fair comparison
cc_idx <- complete.cases(cohort[, lab_predictors])
cat("Complete cases for lab model:", sum(cc_idx),
    sprintf("(%.1f%% of cohort)\n", 100 * mean(cc_idx)))

# Refit base on same subset
mod_base_cc <- glm(
  died ~ age + gender + diabetes + renal + cancer +
    weekend_admit + factor(fiscal_year) + factor(hospital_num),
  data = cohort[cc_idx, ], family = binomial
)

mod_lab <- glm(
  lab_formula,
  data = cohort[cc_idx, ], family = binomial
)
```

### Likelihood ratio test (nested model comparison)

```{r lrt}
lrt <- anova(mod_base_cc, mod_lab, test = "Chisq")

tibble(
  Model     = c("Base", "Base + Labs"),
  Deviance  = round(c(deviance(mod_base_cc), deviance(mod_lab)), 1),
  df        = c(mod_base_cc$df.residual, mod_lab$df.residual),
  `LRT p`   = c(NA, format.pval(lrt$`Pr(>Chi)`[2], digits = 3, eps = 0.001))
) %>%
  kable(caption = "Likelihood ratio test — Base vs Lab-augmented model") %>%
  kable_styling(full_width = FALSE)
```

### AUC comparison (DeLong test)

```{r auc-comparison, fig.height=5, fig.width=6}
pred_base_cc <- predict(mod_base_cc, type = "response")
pred_lab     <- predict(mod_lab, type = "response")

roc_base_cc <- roc(cohort$died[cc_idx], pred_base_cc, quiet = TRUE)
roc_lab     <- roc(cohort$died[cc_idx], pred_lab, quiet = TRUE)

# DeLong test for difference in AUCs
delong <- roc.test(roc_base_cc, roc_lab, method = "delong")

# Side-by-side ROC
ggroc(list(Base = roc_base_cc, `Base + Labs` = roc_lab), linewidth = 0.7) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Base" = "#264653", "Base + Labs" = "#E76F51")) +
  labs(
    title = "ROC comparison — Base vs Lab-augmented",
    subtitle = sprintf(
      "AUC Base: %.3f | AUC +Labs: %.3f | DeLong p = %s",
      auc(roc_base_cc), auc(roc_lab),
      format.pval(delong$p.value, digits = 3, eps = 0.001)),
    x = "Specificity", y = "Sensitivity", colour = "Model"
  ) +
  coord_equal() +
  theme(legend.position = "bottom")
```

### Lab ORs

```{r lab-or-table}
tidy(mod_lab, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term %in% lab_predictors) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 2)),
         p.value = format.pval(p.value, digits = 2, eps = 0.001)) %>%
  select(Lab = term, OR = estimate,
         `2.5%` = conf.low, `97.5%` = conf.high,
         p = p.value) %>%
  kable(caption = "Adjusted ORs for lab predictors (augmented model)") %>%
  kable_styling(full_width = FALSE)
```

---

# 6 — Summary & Limitations

This analysis examined **`r format(nrow(cohort), big.mark = ",")`** adult medical
admissions across **`r n_distinct(cohort$hospital_num)`** GEMINI hospitals (April
2019 – March 2022). **`r format(sum(cohort$died), big.mark = ",")`** in-hospital
deaths occurred (`r round(100 * mean(cohort$died), 1)`%).

**Key findings:**

- Older age, renal disease, and cancer were independently associated with higher
  in-hospital mortality after adjusting for demographics, comorbidities, admission
  timing, and hospital.
- The base model achieved an AUC of **`r round(auc_base[2], 3)`**.
- Addition of early lab values [improved / did not improve] discrimination
  (DeLong p = `r format.pval(delong$p.value, digits = 3)`).

**Limitations:**

1. **Residual confounding** — administrative data cannot capture functional status,
   goals-of-care designation, or illness acuity at presentation.
2. **Simplified comorbidity flags** — our three-condition approach does not replace
   a validated index (Charlson, Elixhauser). A formal comorbidity score should be
   used in a full analysis.
3. **Complete-case lab analysis** — missingness was not random across hospitals.
   Multiple imputation (e.g., `mice`) would better handle this in a full study.
4. **Fixed-effect hospital adjustment** — with many hospitals, a mixed-effects model
   (random intercept for hospital) would be more efficient and generalizable.
5. **No external validation** — model performance may not generalise beyond the
   GEMINI network.

**Next steps:** Restricted cubic splines for age, multiple imputation for lab
missingness, LASSO variable selection across a broader lab panel, and temporal
validation (train on FY 2019–2020, test on FY 2021).
