---
title: "Template — Clinical EHR Cohort Analysis"
author: "Nasim"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: kate
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)

pacman::p_load(
  tidyverse, lubridate, table1, broom,
  janitor, knitr, kableExtra, stringr, pROC
)
```

<!--
==========================================================================
SECTION 1 — COHORT CONSTRUCTION
==========================================================================
QUESTION:  Who is in the study and how did we arrive at the analytic sample?
WHY:       Every observational study must define a reproducible cohort with
           explicit inclusion/exclusion criteria. The attrition table lets
           reviewers assess generalizability and whether exclusions introduced
           selection bias. This is the first thing GEMINI collaborators check.
========================================================================== -->

# 1 — Cohort Construction

```{r load-data}
# input_admin  <- "ipadmdad.csv"
# input_dt_admit <- "admission_date_time"
# input_dt_disch <- "discharge_date_time"

input_raw <- read.csv("INPUT_ADMIN_FILE.csv") %>%
  mutate(
    INPUT_DT_ADMIT = ymd_hm(INPUT_DT_ADMIT),
    INPUT_DT_DISCH = ymd_hm(INPUT_DT_DISCH)
  )
```

We apply sequential inclusion criteria and track attrition at each step.

```{r cohort-exclusions}
n0 <- nrow(input_raw)

# Step 1: Restrict to adults
# input_age <- "age"
cohort <- input_raw %>%
  filter(!is.na(INPUT_AGE), INPUT_AGE >= 18)
n1 <- nrow(cohort)

# Step 2: Non-missing outcome proxy
# input_outcome_raw <- "discharge_disposition"
cohort <- cohort %>%
  filter(!is.na(INPUT_OUTCOME_RAW),
         trimws(as.character(INPUT_OUTCOME_RAW)) != "")
n2 <- nrow(cohort)

# Step 3: Minimum length-of-stay (removes same-day transfers / errors)
cohort <- cohort %>%
  mutate(
    los_hrs = as.numeric(
      difftime(INPUT_DT_DISCH, INPUT_DT_ADMIT, units = "hours"))
  ) %>%
  filter(!is.na(los_hrs), los_hrs >= 4)
n3 <- nrow(cohort)

# Step 4: Study period window
# input_start <- "2019-04-01"
# input_end   <- "2022-04-01"
cohort <- cohort %>%
  filter(INPUT_DT_ADMIT >= ymd("INPUT_START_DATE"),
         INPUT_DT_ADMIT <  ymd("INPUT_END_DATE"))
n4 <- nrow(cohort)

# Attrition table
tibble(
  Step    = c("Initial admissions",
              "Age >= 18",
              "Non-missing discharge disposition",
              "LOS >= 4 hours",
              "Study period (INPUT_START — INPUT_END)"),
  N       = c(n0, n1, n2, n3, n4),
  Dropped = c(NA_integer_, n0 - n1, n1 - n2, n2 - n3, n3 - n4)
) %>%
  kable(caption = "Table S1. Cohort attrition") %>%
  kable_styling(full_width = FALSE)
```

The final analytic cohort includes **`r n4`** encounters.

---

<!--
==========================================================================
SECTION 2 — FEATURE ENGINEERING
==========================================================================
QUESTION:  What variables do we need to derive from the raw data, and how?
WHY:       Raw EHR data is never analysis-ready. Outcomes must be coded from
           disposition codes, continuous variables may need categorisation for
           clinical interpretability, and comorbidities must be derived from
           diagnosis tables using standard code sets (ICD-10-CA). Documenting
           every derivation ensures reproducibility and transparency about
           analytic decisions.
========================================================================== -->

# 2 — Feature Engineering

## 2.1 Outcome: binary endpoint

```{r derive-outcome}
# CIHI discharge disposition codes indicating death: 7, 72, 73
# input_death_codes <- c(7, 72, 73)

cohort <- cohort %>%
  mutate(died = as.integer(INPUT_OUTCOME_RAW %in% c(INPUT_DEATH_CODES)))

cat("Events:", sum(cohort$died),
    sprintf("(%.1f%%)", 100 * mean(cohort$died)))
```

## 2.2 Length of stay

```{r derive-los}
cohort <- cohort %>%
  mutate(
    los_days     = los_hrs / 24,
    los_days_cat = cut(los_days,
                       breaks = c(0, 3, 7, 14, Inf),
                       labels = c("<3d", "3-7d", "7-14d", ">14d"),
                       right  = TRUE)
  )
```

## 2.3 Comorbidity flags from diagnosis codes

We derive binary comorbidity indicators from admission diagnoses
(types M, 1, 2, 3) using ICD-10-CA prefixes. This is **not** a validated
Charlson index — it is a simplified disease-burden proxy for this analysis.

```{r derive-comorbidities}
# input_dx_file <- "ipdiagnosis.csv"
# input_id      <- "genc_id"
# input_dx_code <- "diagnosis_code"
# input_dx_type <- "diagnosis_type"

input_dx_raw <- read.csv("INPUT_DX_FILE.csv")

admit_dx <- input_dx_raw %>%
  filter(INPUT_ID %in% cohort$INPUT_ID,
         INPUT_DX_TYPE %in% c("M", "1", "2", "3"),
         !is.na(INPUT_DX_CODE), trimws(INPUT_DX_CODE) != "")

# Flag each condition per encounter using ICD-10-CA regex
# input_comorb1_pattern <- "E1[01]"   # diabetes
# input_comorb2_pattern <- "N1[78]"   # CKD
# input_comorb3_pattern <- "C"        # cancer

comorb <- admit_dx %>%
  mutate(
    comorb1 = as.integer(str_starts(INPUT_DX_CODE, "INPUT_PATTERN_1")),
    comorb2 = as.integer(str_starts(INPUT_DX_CODE, "INPUT_PATTERN_2")),
    comorb3 = as.integer(str_starts(INPUT_DX_CODE, "INPUT_PATTERN_3"))
  ) %>%
  group_by(INPUT_ID) %>%
  summarise(
    comorb1 = max(comorb1),
    comorb2 = max(comorb2),
    comorb3 = max(comorb3),
    .groups = "drop"
  )

cohort <- cohort %>%
  left_join(comorb, by = "INPUT_ID") %>%
  mutate(across(c(comorb1, comorb2, comorb3), ~ replace_na(.x, 0L)),
         dx_burden = comorb1 + comorb2 + comorb3)
```

```{r comorbidity-summary}
cohort %>%
  summarise(across(c(comorb1, comorb2, comorb3),
                   list(n = sum, pct = ~ round(100 * mean(.x), 1)))) %>%
  pivot_longer(everything(),
               names_to = c("condition", ".value"),
               names_pattern = "(.+)_(n|pct)") %>%
  kable(col.names = c("Condition", "N", "%"),
        caption   = "Comorbidity prevalence in final cohort") %>%
  kable_styling(full_width = FALSE)
```

## 2.4 Admission timing

```{r admission-timing}
cohort <- cohort %>%
  mutate(
    weekend_admit = as.integer(wday(INPUT_DT_ADMIT) %in% c(1, 7)),
    fiscal_year   = if_else(month(INPUT_DT_ADMIT) >= 4,
                            year(INPUT_DT_ADMIT),
                            year(INPUT_DT_ADMIT) - 1L)
  )

cohort %>%
  count(fiscal_year, weekend_admit) %>%
  group_by(fiscal_year) %>%
  mutate(label = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) %>%
  ungroup() %>%
  mutate(weekend_admit = if_else(weekend_admit == 1L, "Weekend", "Weekday")) %>%
  select(weekend_admit, fiscal_year, label) %>%
  pivot_wider(names_from = fiscal_year, values_from = label) %>%
  kable(caption = "Admissions by day type and fiscal year") %>%
  kable_styling(full_width = FALSE)
```

---

<!--
==========================================================================
SECTION 3 — DESCRIPTIVE ANALYSIS
==========================================================================
QUESTION:  How do patient characteristics differ between those who experienced
           the outcome and those who did not?
WHY:       Table 1 is the foundation of every clinical manuscript — it lets
           collaborators verify cohort face-validity, spot data issues (e.g.,
           implausible age ranges), and identify potential confounders before
           modelling begins. It also drives hypothesis generation and informs
           variable selection for the regression.
========================================================================== -->

# 3 — Descriptive Analysis

```{r table1}
cohort_tbl <- cohort %>%
  mutate(
    Outcome  = factor(died, 0:1, c("Survived", "Died")),
    # input_sex <- "gender"
    Gender   = factor(INPUT_SEX, c("M", "F"), c("Male", "Female")),
    Comorb1  = factor(comorb1, 0:1, c("No", "Yes")),
    Comorb2  = factor(comorb2, 0:1, c("No", "Yes")),
    Comorb3  = factor(comorb3, 0:1, c("No", "Yes"))
  )

table1(~ INPUT_AGE + Gender + los_days + Comorb1 + Comorb2 + Comorb3 | Outcome,
       data    = cohort_tbl,
       caption = "Table 1. Patient characteristics by outcome.")
```

Patients who died were older and had a higher prevalence of renal disease and cancer
compared to survivors. Weekend admission proportions appeared similar across groups.

---

<!--
==========================================================================
SECTION 4 — LOGISTIC REGRESSION
==========================================================================
QUESTION:  Which patient-level factors are independently associated with the
           outcome after adjusting for confounders?
WHY:       Multivariable logistic regression isolates the effect of each
           predictor while holding others constant, producing adjusted ORs
           that are the standard currency in clinical epidemiology.
           Hospital is included as a fixed effect to absorb between-site
           variation in case-mix and practice patterns.
           A forest plot communicates results faster than a table for
           clinician audiences.
========================================================================== -->

# 4 — Logistic Regression

## 4.1 Model fitting

```{r logistic-model}
# input_hospital <- "hospital_num"

model <- glm(
  died ~ INPUT_AGE + INPUT_SEX + comorb1 + comorb2 + comorb3 +
    weekend_admit + factor(fiscal_year) + factor(INPUT_HOSPITAL),
  data   = cohort,
  family = binomial
)
```

```{r or-table}
tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(!str_starts(term, "factor\\(INPUT_HOSPITAL\\)")) %>%
  mutate(across(c(estimate, conf.low, conf.high), ~ round(.x, 2)),
         p.value = format.pval(p.value, digits = 2, eps = 0.001)) %>%
  select(Term = term, OR = estimate, `2.5%` = conf.low,
         `97.5%` = conf.high, p = p.value) %>%
  kable(caption = "Table 2. Adjusted ORs (hospital effects suppressed)") %>%
  kable_styling(full_width = FALSE)
```

```{r model-interpretation, results='asis'}
or_main <- tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term == "INPUT_AGE")  # swap for any predictor of interest

cat(sprintf(
  "Each additional year of age was associated with **%.0f%% higher odds** of in-hospital mortality (OR %.2f, 95%% CI %.2f--%.2f, p %s), after adjustment for sex, comorbidities, admission timing, fiscal year, and hospital.",
  (or_main$estimate - 1) * 100,
  or_main$estimate, or_main$conf.low, or_main$conf.high,
  format.pval(or_main$p.value, digits = 2, eps = 0.001)
))
```

## 4.2 Discrimination

```{r model-auc}
cohort$pred_prob <- predict(model, type = "response")
roc_obj <- roc(cohort$died, cohort$pred_prob, quiet = TRUE)
auc_ci  <- ci.auc(roc_obj)

cat(sprintf("AUC: %.3f (95%% CI %.3f–%.3f)",
            auc_ci[2], auc_ci[1], auc_ci[3]))
```

## 4.3 Forest plot

```{r forest-plot, fig.width=8, fig.height=5}
or_df <- tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)",
         !str_starts(term, "factor\\(INPUT_HOSPITAL\\)")) %>%
  mutate(
    term = case_match(
      term,
      "INPUT_AGE"                ~ "Age (per year)",
      "INPUT_SEXM"               ~ "Male sex",
      "comorb1"                  ~ "Comorbidity 1",
      "comorb2"                  ~ "Comorbidity 2",
      "comorb3"                  ~ "Comorbidity 3",
      "weekend_admit"            ~ "Weekend admission",
      "factor(fiscal_year)2020"  ~ "FY 2020 vs ref",
      "factor(fiscal_year)2021"  ~ "FY 2021 vs ref"
    ),
    significant = p.value < 0.05
  )

ggplot(or_df, aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high, colour = significant),
    size = 0.6
  ) +
  scale_colour_manual(
    values = c("TRUE" = "#D62828", "FALSE" = "#264653"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p >= 0.05"),
    name   = NULL
  ) +
  labs(
    x       = "Adjusted Odds Ratio (95% CI)",
    y       = NULL,
    title   = "Adjusted Predictors of In-Hospital Mortality",
    caption = "Hospital fixed effects estimated but suppressed."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

---

<!--
==========================================================================
SECTION 5 — EARLY LAB VALUES (FIRST 24 h)
==========================================================================
QUESTION:  Do early physiological measurements (lab results collected within
           24 h of admission) add predictive value beyond demographics and
           diagnosis codes?
WHY:       Administrative data captures who the patient is but not how sick
           they are at presentation. Lab values are the closest proxy to
           acute severity in structured EHR data. Showing you can link
           time-stamped clinical data to encounters and handle missingness
           is a core GEMINI data-science competency.
========================================================================== -->

# 5 — Early Lab Values (First 24 h)

```{r lab-features}
# input_lab_file <- "lab.csv"
# input_lab_dt   <- "collection_date_time"
# input_lab_test <- "test_type_mapped_omop"
# input_lab_val  <- "result_value"

input_lab_raw <- read.csv("INPUT_LAB_FILE.csv")

lab_early <- input_lab_raw %>%
  filter(INPUT_ID %in% cohort$INPUT_ID) %>%
  inner_join(cohort %>% select(INPUT_ID, INPUT_DT_ADMIT), by = "INPUT_ID") %>%
  mutate(
    INPUT_LAB_DT    = ymd_hm(INPUT_LAB_DT),
    hrs_since_admit = as.numeric(
      difftime(INPUT_LAB_DT, INPUT_DT_ADMIT, units = "hours"))
  ) %>%
  filter(hrs_since_admit >= 0, hrs_since_admit <= 24)

cat("Lab records within first 24 h:", nrow(lab_early))
```

```{r lab-summary}
lab_summary <- lab_early %>%
  group_by(INPUT_ID, INPUT_LAB_TEST) %>%
  summarise(
    first_value = INPUT_LAB_VAL[which.min(INPUT_LAB_DT)],
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = INPUT_LAB_TEST, values_from = first_value)

cohort <- cohort %>%
  left_join(lab_summary, by = "INPUT_ID")
```

```{r lab-missingness}
# Quick missingness check on key lab columns
# Adjust column names to whatever labs are present in the data
lab_cols <- names(lab_summary)[-1]  # drop ID column

cohort %>%
  summarise(across(all_of(lab_cols), ~ round(100 * mean(is.na(.x)), 1))) %>%
  pivot_longer(everything(), names_to = "Lab test", values_to = "% missing") %>%
  arrange(desc(`% missing`)) %>%
  kable(caption = "Lab missingness (first 24 h value)") %>%
  kable_styling(full_width = FALSE)
```

---

# 6 — Summary

This analysis identified **`r sum(cohort$died)`** in-hospital deaths
(`r round(100 * mean(cohort$died), 1)`%) among **`r nrow(cohort)`** adult medical
admissions across GEMINI hospitals during the study window. Older age,
comorbidity burden, and [condition] were associated with higher mortality.
The model achieved an AUC of **`r round(auc_ci[2], 3)`**. Next steps include
assessment of non-linearity in age (splines), incorporation of early lab
values as predictors, and evaluation of model calibration.
